### 2022-02-22 doc/university-covid.tex
done?

Mathematical models have played a crucial role in guiding pandemic responses, especially for reopening university campuses.
Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions in Princeton University.
We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited;
mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
Predicting future outbreaks and the effectiveness of intervention measures on campus is expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.

@@

Mathematical models have been particularly useful in planning campus reopenings.
Researchers from various institutions---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
Coupling modeling efforts with real-life implementations in university campuses further provided unique opportunities to directly test model-based predictions of intervention effects in preventing the transmission of SARS-CoV-2 \citep{frazier2022modeling}---
each university campus offers a relatively well-controlled epidemic setting with a behaviorally homogeneous population (especially among undergraduate students).
Mass asymptomatic testing also provides near-perfect ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.

University campuses also present unique challenges to controlling an outbreak.
High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
These kinds of contacts are inherently difficult to trace, making contact tracing less effective.
The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University during the same time period despite similar levels of testing and vaccination.
Here, we choose to focus on the spread of SARS-CoV-2 on Princeton University campus alone to eliminate heterogeneities inherent to such comparisons.

In this study, we analyze the spread of SARS-CoV-2 on Princeton University campus (\fref{princeton}).
We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
Princeton University is located in Mercer County, New Jersey, USA and consists of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
Throughout the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Asymptomatic individuals who tested positive were required to isolate for 10 days.
Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
PCR positives were exempt from asymptomatic testing for 90 days.
Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
All data used in this analysis are publicly available on the Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University.}
(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
Red lines represent the weekly number of cases in Mercer County.
Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
(D--F) Correlations between the weekly number of cases in Princeton University and in Mercer County.
Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
(G) Ratios between weekly numbers of cases reported from Princeton University and Mercer County.
Solid lines represent the observed ratios.
Dotted lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
\label{fig:princeton}
}
\end{figure}

@@

In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in Princeton University also coincided with the decrease in the number of cases in Mercer County.
The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).

@@

Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from Princeton University and those from Mercer county:
fall 2020 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}D); spring 2020 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}E); and fall 2021 ($\rho = 0.90$ (95\% CI: 0.78--0.96; $p < 0.001$), \fref{princeton}F). 
These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
We also find strong correlations between the weekly logged numbers of cases from Princeton University and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).

@@

Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in Princeton University during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
We also find that some variations in the ratio around the mean can be explained by holiday effects.
For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.

\section*{Mathematical model}

We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the Princeton University campus.
This model was initially developed and used throughout the pandemic to inform policy decisions in Princeton University, including the frequency of asymptomatic tests and the number of isolation beds required.
We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in Princeton University without an over-proliferation of parameters.
The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.

Infection processes are modeled based on standard compartmental structures.
Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
Recovered individuals are assumed to be immune to reinfections throughout a semester.
Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}.
More specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.

Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
here, infectious contacts refer to contacts that would result in infection when the contactee is susceptible to infection.
The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively. Pre-symptomatic and (a)symptomatic reproduction numbers are calculated based on the assumed value of the proportion of presymptomatic transmission $p_p = 0.5$: $\mathcal R_p/\mathcal R_s = p_p/(1 - p_p)$.
On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events.
Infectious contacts from the community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, contactees have a reduced probability of infection corresponding to their susceptibility (discussed later).

All individuals on campus are assumed follow a pre-determined asymptomatic testing plan at a fixed frequency---
for example, under weekly testing, one individual can get sampled on days 1, 8, 15, and so forth, while another individual get sampled on days 2, 9, 16, and so forth.
We assume that test results come back after one day.
Symptomatic individuals can choose to take rapid PCR tests (with results returning on the same day) with a given probability on each day until their symptoms resolve---this probability is set to 1 for simulations presented in the main text.
We further assume that symptomatic individuals are isolated immediately when they submit their samples until they receive negative results.
All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days.
Isolated individuals are assumed to no longer transmit infections.
We assume that PCR tests have 95\% sensitivity and 100\% specificity.

@@

In this study, we use this model to retrospectively analyze past outbreaks and further predict future epidemic trajectories for the Omicron variant.
First, we try to match our model to epidemic patterns seen on campus by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume that all other parameters are equal across different groups for simplicity. 
We further assume that the population mixes homogeneously.
In practice, epidemiological parameters, such as contact rates and the proportion of asymptomatic infections, likely differ across groups, and mixing patterns likely exhibit assortativity (e.g., undergraduate students are more likely to infect undergraduate students).
Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation.pdf}
\caption{
\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University.}
(A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community transmission $\theta$.
For each parameter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases. 
Heat maps represent medians of the logged sum of squared differences.
(B, E, H) Model predictions. 
Solid lines represent median predictions.
Shaded areas represent 90\% quantiles for the best matching parameter set.
Points represent the observed data.
(C, F, I) Correlations between model predictions with observed data.
Colored solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
Dashed lines represent the one-to-one line.
\label{fig:matching}
}
\end{figure}

For fall 2020, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
We find that a low level of contacts $\Rc=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of a sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S6).
Distancing measures on campus and contact tracing efforts likely contributed to lowering contact levels $\Rc$.

@@

A similar set of parameters can capture the observed dynamics in spring 2020.
The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S7). 
Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
We also find that initial infections are required to match relatively high levels of cases in the beginning of semester (\fref{matching}E). 
Once again, the predicted and the observed numbers of cases are positively correlated ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity.pdf}
\caption{
\textbf{Sensitivity of the spread of Omicron on campus to underlying epidemiological and intervention conditions.}
(A, C, E, G, I, K) Predicted numbers of Omicron infections per week on campus. 
Colored lines represent median predictions.
Shaded areas represent 90\% quantiles across 100 simulations.
We vary one parameter at a time while holding all other parameters constant at their nominal values.
(B, D, F, H, J, L) Predicted numbers of total Omicron infections on campus.
All other parameters are the same as in Figure 2.
\label{fig:sensitivity}
}
\end{figure}

>>>>

Mathematical models have played a crucial role in guiding pandemic responses; university campuses provide a particularly well documented case for institutional outbreaks.
Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions for the Princeton University campus.
We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic.
Mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
Predicting future outbreaks, and the effectiveness of intervention measures on campus, is then expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.

@@

Mathematical models have also been widely deployed in planning campus reopenings.
Researchers from various institutions in the US---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
Coupling modeling efforts with real-life implementations in university campuses further provided unique opportunities to directly test model-based predictions of intervention effects in preventing the transmission of SARS-CoV-2 \citep{frazier2022modeling}---
each university campus offers a relatively well-controlled epidemic setting with a behaviorally homogeneous population (especially among undergraduate students).
Mass asymptomatic testing also provides near-perfect ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.

University campuses also present unique challenges to controlling an outbreak.
High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
These kinds of contacts are inherently difficult to trace, making contact tracing less effective.
The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period despite similar levels of testing and vaccination.
Here, we choose to focus on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.

In this study, we analyze the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
PU is located in Mercer County, New Jersey, USA;
the population comprises of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
Throughout the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Asymptomatic individuals who tested positive were required to isolate for 10 days.
Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
PCR positives were exempt from asymptomatic testing for 90 days.
Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University (PU).}
(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
Red lines represent the weekly number of cases in Mercer County.
Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
(D--F) Correlations between the weekly number of cases in PU and in Mercer County.
Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
(G) Ratios between weekly numbers of cases reported from PU and Mercer County.
Solid lines represent the observed ratios.
Dotted lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
\label{fig:princeton}
}
\end{figure}

@@

In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).

@@

Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
fall 2020 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}D); spring 2020 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}E); and fall 2021 ($\rho = 0.90$ (95\% CI: 0.78--0.96; $p < 0.001$), \fref{princeton}F). 
These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).

@@

Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
We also find that some variations in the ratio around the mean can be explained by holiday effects.
For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.

\section*{Mathematical modeling of past outbreaks}

We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the PU campus.
This model was initially developed and used throughout the pandemic to inform policy decisions in PU, including the frequency of asymptomatic tests and the number of isolation beds required.
We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in PU without an over-proliferation of parameters.
The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.

Infection processes are modeled based on standard compartmental structures (Supplementary Figure S6).
Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
Recovered individuals are assumed to be immune to reinfections throughout a semester.
Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
Assumed parameters are broadly consistent with other modeling approaches  \citep{brett2020transmission,lavezzo2020suppression}.

Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
here, infectious contacts refer to contacts that would result in infection when the contactee is susceptible to infection.
The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively. Pre-symptomatic and (a)symptomatic reproduction numbers are calculated based on the assumed value of the proportion of presymptomatic transmission $p_p = 0.5$: $\mathcal R_p/\mathcal R_s = p_p/(1 - p_p)$.
On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
Infectious contacts from the community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, contactees have a reduced probability of infection corresponding to their susceptibility (discussed later).

All individuals on campus are assumed follow a pre-determined asymptomatic testing plan at a fixed frequency---
for example, under weekly testing, one individual can get sampled on days 1, 8, 15, and so forth, while another individual get sampled on days 2, 9, 16, and so forth.
We assume that test results come back after one day.
Symptomatic individuals can choose to take rapid PCR tests (with results returning on the same day) with a given probability on each day until their symptoms resolve---this probability is set to 1 for simulations presented in the main text.
We further assume that symptomatic individuals are isolated immediately when they submit their samples until they receive negative results.
All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days.
Isolated individuals are assumed to no longer transmit infections.
We assume that PCR tests can detect infections from individuals who are in pre-symptomatic, symptomatic, and asymptomatic stages with 95\% sensitivity and 100\% specificity.

@@

In this study, we use this model to retrospectively analyze past outbreaks and further predict future epidemic trajectories for the Omicron variant.
First, we try to match our model to epidemic patterns seen on campus by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume that all other parameters are equal across different groups for simplicity. 
We further assume that the population mixes homogeneously.
While these assumptions are most conservative, epidemiological parameters  and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation.pdf}
\caption{
\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University (PU).}
(A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community transmission $\theta$.
For each parameter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases. 
Heat maps represent medians of the logged sum of squared differences.
(B, E, H) Model predictions. 
Solid lines represent median predictions.
Shaded areas represent 90\% quantiles for the best matching parameter set.
Points represent the observed data.
(C, F, I) Correlations between model predictions with observed data.
Colored solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
Dashed lines represent the one-to-one line.
\label{fig:matching}
}
\end{figure}

For fall 2020, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
We find that a low level of contacts $\Rc=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of a sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S7).
Distancing measures on campus and contact tracing efforts likely contributed to lowering contact levels $\Rc$.

@@

A similar set of parameters can capture the observed dynamics in spring 2020.
The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S8). 
Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
We also find that initial infections are required to match relatively high levels of cases in the beginning of semester (\fref{matching}E). 
Once again, the predicted and the observed numbers of cases are positively correlated ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity.pdf}
\caption{
\textbf{Sensitivity of the spread of Omicron on campus to underlying epidemiological and intervention conditions.}
We perform sensitivity analysis for potential Omicron outbreaks on campus by varying contact reproduction number $\Rc$ (A--B), asymptomatic testing frequency (C--D), vaccination strategy (E--F), community contacts (G--H), prevailing immunity (I--J), and isolation strategy (K--L).
(A, C, E, G, I, K) Predicted numbers of Omicron infections per week on campus. 
Colored lines represent median predictions.
Shaded areas represent 90\% quantiles across 100 simulations.
We vary one parameter at a time while holding all other parameters constant at their nominal values.
We present representative simulations in these panels; all other simulations can be found in Supplementary Figures S11--S14.
(B, D, F, H, J, L) Predicted numbers of total Omicron infections on campus.
All other parameters are the same as in Figure 2.
\label{fig:sensitivity}
}
\end{figure}
